---
title: "Coursera_MachineLearningFinalAssignment"
author: "Collier King"
date: "June 12, 2016"
output: html_document
---
# Practical Machine Learning - Final Project

## Introduction & Background

The rise of the internet of things has created an entire industry of personal activity
tracking devices. These devices allow the user to measure their activities and seek ways to 
improve their health. In addition to merely measuring the level of activity a user engages in,
it might be useful to the user to measure how well they are actually performing the activities.

In this analysis, we will be using the data from Weight Lifting Exercises dataset. 
The data consist of results for different exercises as gathered by accelerometers on the belt, 
forearm, and dumbells of six study subjects.  Our data are separated into training and testing sets.  
The training set holds an indicator column for the quality of the exercise performed, 
the testing set does not have this column.  
This analysis will attempt to correctly predict this indicator column on the test set.

```{r, include=FALSE}
#Load packages
library(caret)
setwd("C:\\Users\\Collier\\Dropbox\\Skills\\R\\Case Practice\\Machine Learning\\Coursera_MLFinalProject\\Coursera_MLFinalProject")
```

## Data Preparation
```{r, message=FALSE}
dtrain <- read.csv(file="pml-training.csv", header=TRUE, as.is = TRUE, stringsAsFactors = FALSE, sep=',', na.strings=c('NA','','#DIV/0!'))
dtest <- read.csv(file="pml-testing.csv", header=TRUE, as.is = TRUE, stringsAsFactors = FALSE, sep=',', na.strings=c('NA','','#DIV/0!'))

# Convert response variable "classe" to factor type
dtrain$classe <- as.factor(dtrain$classe)

#Remove missing data
NAindex <- apply(dtrain,2,function(x) {sum(is.na(x))}) 
dtrain <- dtrain[,which(NAindex == 0)]

NAindex <- apply(dtest,2,function(x) {sum(is.na(x))}) 
dtest <- dtest[,which(NAindex == 0)]

```

## Pre-processing and Scaling
The prediction variables are preprocessed by K-nearest neighbors imputation, centering and scaling.
```{r, message=FALSE}
var <- which(lapply(dtrain, class) %in% "numeric")

preObj <-preProcess(dtrain[,var],method=c('knnImpute', 'center', 'scale'))

dtrainscale <- predict(preObj, dtrain[,var])
dtrainscale$classe <- dtrain$classe

dtestscale <-predict(preObj,dtrain[,var])
```

## Remove variables with near zero variance
Features are reduced by removing variables with near zero variance.  This variables will have low importance in the prediction process.
```{r, message=FALSE}
var0 <- nearZeroVar(dtrainscale,saveMetrics = TRUE)
dtrainscale <- dtrainscale[,var0$nzv==FALSE]

var0 <- nearZeroVar(dtestscale,saveMetrics = TRUE)
dtestscale <- dtestscale[,var0$nzv==FALSE]

```

## Create Cross-Validation Set
The training set it then divided into two parts (60/40) for training and cross validation, respectively.
```{r, message=FALSE}
set.seed(5424613)
inTrain <- createDataPartition(dtrainscale$classe, p = 0.6, list = FALSE)
train <- dtrainscale[inTrain,]
xVal <- dtrainscale[-inTrain,]
```

## Model Building on Training Set
I chose to use a random forest model. 27 predictor variables will be used with 3-fold cross validation for the train control method.
```{r, message=FALSE}
startTime <- Sys.time();

modRF <- train(classe ~., method="rf", data=train, trControl=trainControl(method='cv'), number=3, allowParallel=TRUE)

endTime <- Sys.time()
endTime - startTime

modRF
```

## Measuring Model Accuracy on Training Dataset
Initial model prediction executed on training Dataset.
```{r, message=FALSE}
trainPred <- predict(modRF,train)
confusionMatrix(trainPred,train$classe)
```

## Measuring Accuracy on Cross Validation Dataset
Cross validation dataset testing of model. The model's accuracy is 99.17% with an out of sample error of .83%.
```{r, message=FALSE}
xValPred <- predict(modRF,xVal)
confusionMatrix(xValPred,xVal$classe)
```

## Model Evaluation on Testing Dataset
Generate predictions on actual test set data. Write prediction files for submission.
```{r, message=FALSE}
testPred <- predict(modRF,dtest)

# Write prediction data to file
path = "./Predictions"
MLpred_submit = function(x) {
    n = length(x)
    for(i in 1: n) {
        filename = paste0("problem_id_", i, ".txt")
        write.table(x[i], file=file.path(path, filename), 
                    quote=FALSE, row.names=FALSE, col.names=FALSE)
    }
}
MLpred_submit(testPred)
```